Magnetic Phase Transitions in an Aperiodic Monotile

### 2023-06-09 paper/main.tex
Working on paper

\begin{abstract}
  A shape known as the ``einstein tile'' has been discovered which tiles the plane aperiodically. We define the Ising, XY, and transverse Ising models on the einstein tiling and compute the magnetization and susceptibility of the tiling as a function of temperature using Monte Carlo methods. This allows us to produce phase diagrams of the relationship between the critical temperature of the tiling and the dimensions of the einstein tile.
\end{abstract}

@@

\section{Phase Transitions and Spin Models}
The Ising model was one of the first statistical mechanics models to contain a phase transition, and remains one of the only exactly solvable examples. It consists of defining a variable $s_i = \pm 1$ at every site of a lattice with $N$ sites as a classical approximation of the up-or-down spin of a fermion. On further analogy with fermionic spins, it asserts that the spins interact by defining the classical Hamiltonian
\begin{equation}
  H = -J \sum_{\braket{ij}} s_i s_j + h \sum_i s_i
  \label{eqn:ising}
\end{equation}
where $h$ is an applied magnetic field. This model's phase transition is most visible in the magnetization, defined as
\begin{equation}
  M = \frac{1}{N}\braket{\sum_{i} s_i}
  \label{eqn:magnetization}
\end{equation}
where the brackets indicate an average over the statistical ensemble of states, weighted by the Boltzmann factor $e^{-\beta H}$. Another interesting parameter is the susceptibility
\begin{equation}
  \chi = \frac{\partial M}{\partial h} = \frac{1}{N} \braket{\sum_{i} (s_i - M)^2}.
\end{equation}
Now that the susceptibility has been defined, we will set $h$ to zero and study the Ising model in the zero magnetic field limit.

The reason why a phase transition occurs is as follows. The Hamiltonian displays $s_i\rightarrow -s_i$ symmetry when when $h=0$. If the thermodynamic system explores all of phase space, then the average thermodynamic state should possess the same symmetries of the Hamiltonian so that $M=0$. However, for low temperature we also expect the system to attain the lowest possible energy. This state consists of all spins up (or all spins down) has $M = 1$ (or $-1$), which violates the expectation of no magnetization. The resolution to this apparent contradiction that there exists a critical temperature $T_c$ below which energetic considerations dominate and the crystal is magnetic ($|M|>0$), and above which the symmetry of phase space dominates and the crystal is not magnetic ($M = 0$).

@@

In addition to the Ising model, we study the XY model which consists of promoting $s_i$ to a two-dimensional unit vector defined at every lattice point. The interaction term in the Hamiltonian becomes $\bm s_i \cdot \bm s_j$. This XY model has different $T_c$ than the Ising model and can even contain vortices, which are topologically protected quasiparticles where the curl of $\bm s_i$ is nonzero. A third spin model is the Transverse Ising Model (TIM), which consists of replacing $s_i$ in Eq.~\ref{eqn:ising} with the Pauli matrices $\bm \sigma_i$ and $\bm \sigma_j$ and interpreting the result as a quantum Hamiltonian. Past research \jtd{citation needed} has shown that the $d$-dimensional TIM is equivalent to a classical Ising model in $d+1$ dimensions in the limit of a very strong interaction strength in the extra dimension. Thus, critical temperatures for all three of these models can be found using the same techniques.

In this final project, we will use Monte Carlo methods discussed in the next section to evaluate the critical temperature of these three models for the recently discovered einstein tiling (figure \ref{fig:einstein}). The einstein tile (German for ``one stone'') is the first shape known to tile the plane in a non-repeating pattern \cite{smith2023aperiodic}. Such aperiodic tilings are called quasicrystals, and a two-tile version known as the Penrose tiling has gained much traction in the literature \jtd{mention its critical points. Also show the einstein tile}. An interesting fact about the einstein tile is that its dimensions are adjustable. The tile consists of two side lengths $a$ and $b$, and any ratio of $a/b$ leads to an aperiodic tiling except $a/b=0$, 1, or $\infty$. If we treat the spin-spin interaction as weaker across the longer bonds, then this $a/b$ ratio represents a free parameter of the crystal, so that the critical temperature is a function of $a/b$. Generally, we take the interaction strength $J$ across a bond of length $\ell$ to be $J = J_0 \ell^{-2}$ where $J_0$ is constant.

@@

\section{Cluster Monte Carlo Methods: the Wolff Algorithm}
Solving the Ising model is generally difficult. Ising himself solved the Ising model for a one-dimensional regular crystal, and a solution was produced 20 years later for a two dimensional square lattice. The critical temperature can sometimes be worked out by a clever trick for the Ising model when the lattice is self-dual, such as the square lattice \jtd{cite}. Even when the lattice is not self-dual, if the lattice shares enough properties with its dual the connection can be used to determine the critical temperature \jtd{cite}.

However, the einstein lattice (like many lattices) satisfies none of these properties, so that we are reduced to numerically computing the ensemble average of Eq.~\ref{eqn:magnetization}. This average is extremely expensive to compute because the number of states in the ensemble is $2^N$ for the Ising model. However, approximate results can be obtained by computing the sum over a much smaller, random selection of states $K$. This approximation technique is called a Monte Carlo method.

@@

\section{Results}

@@

The upper left panel of Figure \ref{fig:square} shows the magnetization $M$ and susceptibility $\chi$ of the Ising, XY, and TI models executed on a square lattice as a function of the lattice width $L = \sqrt{N}$. The critical temperature of the Ising model on this lattice is known to be $T_c = \frac{2}{\ln (1 + \sqrt{2})} \approx 2.27$ in units where $k_B = J = 1$ due to the Onsager solution, and indeed the magnetization falls sharpest near this location. For the XY model, the critical temperature is approximately $0.8816$, which is likewise close to the drop in $M$. The location of these transitions is not perfect, and $M$ it is not strictly zero for $T<T_c$ due to finite-size effects. Larger lattice sizes lead to smaller $M$ for $T<T_c$. In the XY model the critical temperature is substantially lower due to the increased degrees of freedom of the spin system. Furthermore, $M$ approaches its low-temperature limit of 1 linearly for low temperature rather than exponentially, which is a qualitative difference between the Ising and XY models.

\begin{figure*}
  \centering
  \includegraphics[width=0.49\linewidth]{../figs/square.png}\hfill
  \includegraphics[width=0.49\linewidth]{../figs/penrose.png}
  \caption{Magnetization and susceptibility as a function of $T_c$ under the XY and Ising models. \textit{Left}: A square lattice with $N=128^2=16,384$ sites. \textit{Right}: A Penrose lattice with $N=21,316$ sites.}
  \label{fig:square}
\end{figure*}

@@

In both crystals and both models, the point of maximum susceptibility and maximum $dM/dT$ occur at approximately the same location. For the sake of approximating the critical temperature, we take the $T$ point at which the steepest line tangent to the $M(T)$ curve intersects the $M=1$ axis. This approximation results in $T_c\approx 2.26$ for the Ising model on a square lattice and $T_c\approx 0.922$, within a few percent of the true result.

To demonstrate that the $M$-$T$ relationship is not strongly dependent on the lattice, we also compute the same quantities for the Penrose lattice, shown in the right two panels of figure \ref{fig:square}. We see the same behavior in how $M$ and $\chi$ approach $T=0$ and $\infty$. The behavior near $T_c$ is slightly different in the steepness of the curves due to the difference in the number of sites. The major difference between the two is the actual value of $T_c$, which is slightly higher for Penrose. \jtd{Why? State TC and figure out a way to put errors on it? Also draw TC on the plot.}

The square lattice can be stretched into a rectangular lattice by allowing the horizontal and vertical bonds to differ in length. Assigning different interaction strengths $J_1$ and $J_2$ to the two bond lengths gives a free parameter $J_1/J_2$. Figure \ref{fig:ising-rect} shows the critical temperature as a function of $J_1/J_2$ where we have fixed the tile area to be constant: $J_1J_2 = 1$.

A similar procedure shows the phase diagram for the einstein tiling, with the ratio of side lengths $a/b$ as the free parameter

\section{Conclusion}

>>>>

\begin{abstract}
  A family of shapes known as ``einstein tiles'' has been discovered which tile the plane aperiodically. We define the Ising, XY, and Heisenberg models on the einstein tilings and compute the magnetization and susceptibility of the tiling as a function of temperature using Monte Carlo methods. This allows us to produce phase diagrams of the relationship between the critical temperature of the tiling and the dimensions of the einstein tile.
\end{abstract}

@@

\section{Introduction}
The Ising model was one of the first statistical mechanics models to contain a phase transition, and remains one of the only exactly solvable examples. It consists of defining a variable $s_i = \pm 1$ at every site of a lattice with $N$ sites as a classical approximation of the up-or-down spin of a fermion. On further analogy with fermionic spins, it asserts that the spins interact by defining the classical Hamiltonian
\begin{equation}
  H = -J \sum_{\braket{ij}} s_i s_j + h \sum_i s_i
  \label{eqn:ising}
\end{equation}
where $h$ is an applied magnetic field. This model's phase transition is most visible in the magnetization, defined as
\begin{equation}
  M = \frac{1}{N}\braket{\sum_{i} s_i}
  \label{eqn:magnetization}
\end{equation}
where the brackets indicate an average over the statistical ensemble of states, weighted by the Boltzmann factor $e^{-\beta H}$.

The phase transition occurs for the following reason. For $h=0$, the Hamiltonian displays $s_i\rightarrow -s_i$ symmetry. If the thermodynamic system explores all of phase space, then the average thermodynamic state should possess the same symmetries of the Hamiltonian so that $M=0$. However, for low temperature we also expect the system to attain the lowest possible energy. This state consists of all spins up (or all spins down) has $M = 1$ (or $-1$), which violates the expectation of no magnetization. The resolution to this apparent contradiction that there exists a critical temperature $T_c$ below which energetic considerations dominate and the crystal is magnetic ($|M|>0$), and above which the symmetry of phase space dominates and the crystal is not magnetic ($M = 0$).

@@

The solution to a spin model depends on the lattice on which the model is defined. For example, an exact solution exists for the two-dimensional Ising model defined on the square lattice which yields an exact formula for the critical temperature. In the absence of a full solution, a series expansions in $\beta$ (valid at high temperature) or $e^{-\beta}$ (valid at low temperature) can be written for the model's thermodynamical quantities. For self-dual lattices such as the square lattice, the coefficients of the high- and low-temperature expansions are the same --- a phenomenon known as Kramers-Wannier duality \cite{kramers}. This duality leads to an analytical solution for the critical temperature even in the absence of a full solution. Even for lattices which are not self-dual, it is sometimes possible to use a similar technique to relate the series expansions of the lattice and its dual to each other. This can be done for the hexagonal lattice, for example, which is dual to the triangular lattice \cite{bhattacharjee1995fifty}.

In more complicated crystals, however, the only available approaches to computing $T_c$ are numerical. For example, the Penrose lattice consists of two tiles which fill the two-dimensional plane aperiodically, making it difficult to model the system analytically. Ref. \cite{penrose-ising} computes magnetism as a function of temperature using numerical methods and finds that the aperiodic nature of the lattice does not affect the physics at a qualitative level, since $M(T)$ inherits most of its properties from the very general Landau theory.

Recently, the first \textit{single} tile which fills the plane aperiodically has been discovered, and named the einstein tile (German for ``one stone'') (figure \ref{fig:einstein}) \cite{smith2023aperiodic}. An interesting fact about the einstein tile is that its dimensions are adjustable. The tile consists of two side lengths $a$ and $b$, and any ratio of $a/b$ leads to an aperiodic tiling except $a/b=0$, 1, or $\infty$. If we treat the spin-spin interaction as weaker across the longer bonds, then this $a/b$ ratio represents a free parameter of the crystal, so that the critical temperature is a function of $a/b$. The phase diagram of $T_c(a/b)$ is therefore an intrinsic property of the einstein tiling, more so than the commonplace behavior of $M(T)$.

@@

In this final project, we use numerical methods to compute the phase diagram $T_c(a/b)$ for the einstein tile under the Ising model and two similar models. We also compute the phase diagram for a rectangular lattice with bond lengths $a$ and $b$ to compare to the einstein lattice phase diagram, later using this calculation to solve for the one-dimensional transverse Ising model phase transition. The structure of the paper is as follows: in section \ref{sec:spin-models}, we briefly review the spin models. In section \ref{sec:wolff} we discuss the numerical methods used to compute $T_c$. We present the phase diagrams in section \ref{sec:results}, followed by section \ref{sec:tim} on the transverse Ising model. We conclude with section \ref{sec:conclusion}.

\section{Spin models}
\label{sec:spin-models}

In addition to the Ising model Hamiltonian (Eq.~\ref{eqn:ising}), we study the XY model which consists of promoting $s_i$ to a two-dimensional unit vector defined at every lattice point. The interaction term in the Hamiltonian becomes $\bm s_i \cdot \bm s_j$. This XY model has different $T_c$ than the Ising model and can even contain vortices, which are topologically protected quasi-particles where the curl of $\bm s_i$ is nonzero. A third spin model is the Heisenberg model, which is similar to the XY model except that $s_i$ is promoted to a three-dimensional unit vector. Unlike the XY model, the Heisenberg model may not display a true phase transition in the thermodynamic limit \cite{tomita2014finite}. Nevertheless, it has phase-transition-like behavior in that $M$ experiences a large drop in a small temperature range. The Ising, XY, and Heisenberg models have progressively more degrees of freedom, and their critical temperatures are therefore progressively lower.

A fourth spin model is the Transverse Ising Model (TIM), which consists of replacing $s_i$ in Eq.~\ref{eqn:ising} with the Pauli matrices $\bm \sigma_i$ and $\bm \sigma_j$ and interpreting the result as a quantum Hamiltonian. A simple proof using the path integral approach (appendix \ref{app:tim-equivalence}) shows that the $d$-dimensional TIM is equivalent to a classical Ising model in $d+1$ dimensions in the limit of a very strong interaction strength in the extra dimension. Thus, critical temperatures for all four of these models can be found using the same techniques.

Beyond the magnetization $M$, another interesting property of a spin model is its susceptibility $\chi = \frac{dM}{dh}$. Even in a model which sets $h=0$, $\chi$ an be defined as follows by the fluctuation-dissipation theorem (appendix \ref{app:fluctuation-dissipation}):
\begin{equation}
  \chi = \frac{\partial M}{\partial h} = \frac{1}{N}\braket{\sum_i(s_i - M)^2}.
\end{equation}
The susceptibility diverges at the critical temperature and otherwise falls off to zero.

\section{Cluster Monte Carlo Methods: the Wolff Algorithm}
\label{sec:wolff}

To find the critical temperature of the einstein lattice, we must numerically compute the ensemble average of Eq.~\ref{eqn:magnetization}. This average is extremely expensive to compute because the number of states in the ensemble is $2^N$ for the Ising model. However, approximate results can be obtained by computing the sum over a much smaller, random selection of states $K$. This approximation technique is called a Monte Carlo method.

@@

\section{Phase Diagrams}
\label{sec:results}

@@

The upper left panel of Figure \ref{fig:square} shows the magnetization $M$ and susceptibility $\chi$ of the Ising, XY, and Heisenberg models executed on a square lattice as a function of the lattice width $L = \sqrt{N}$. The magnetization drops rapidly near $T_c$, shown as vertical dotted lines. The location of these transitions is not perfect, and $M$ it is not strictly zero for $T<T_c$ due to finite-size effects. Larger lattice sizes lead to smaller $M$ for $T<T_c$. The XY and Heisenberg models have substantially lower $T_c$ due to the increased degrees of freedom of the spin system. Furthermore, $M$ approaches its low-temperature limit of 1 linearly for the Heisenberg and XY models, while the approach is exponential for the Ising model. This is a qualitative difference between the Ising and XY models.

\begin{figure*}
  \centering
  \includegraphics[width=0.49\linewidth]{../figs/square.pdf}\hfill
  \includegraphics[width=0.49\linewidth]{../figs/penrose.pdf}
  \caption{Magnetization and susceptibility as a function of $T_c$ under the XY and Ising models. \textit{Left}: A square lattice with $N=128^2=16,384$ sites. \textit{Right}: A Penrose lattice with $N=21,316$ sites.}
  \label{fig:square}
\end{figure*}

@@

In both crystals and both models, the point of maximum susceptibility and maximum $dM/dT$ occur at approximately the same location. For the sake of approximating the critical temperature, we take the $T$ point at which the steepest line tangent to the $M(T)$ curve intersects the $M=1$ axis. The critical temperatures obtained from this approximation are shown in table \ref{tab:tc}. These temperatures are very close to the true values known for the square lattice: For the Ising model, the Onsager solution gives $T_c = \frac{2}{\ln(1 + \sqrt{2})} \approx 2.2691$, and for the XY model MCMC methods give $T_c \approx 0.8816$. The Heisenberg model does not have a true phase transition, but this fact is difficult to verify in a lattice of finite size, which cannot display true phase transitions until $N\rightarrow \infty$. 
\begin{table}
  \centering
  \begin{tabular}{c|cc}
    \hline \hline
    & Square & Penrose \\ \hline
    Ising & 2.26 & 2.36\\
    XY & 0.928 & 0.943\\
    Heisenberg & 0.493 & 0.462\\ \hline \hline
  \end{tabular}
  \caption{Critical temperatures for the Ising, XY, and Heisenberg models defined on the square and Penrose lattices. These values are very close to the true values (see text).}
  \label{tab:tc}
\end{table}

To demonstrate that the $M(T)$ relationship is not strongly dependent on the lattice, we also compute the same quantities for the Penrose lattice, shown in the right two panels of figure \ref{fig:square}. We see the same behavior in how $M$ and $\chi$ approach $T=0$ and $\infty$. The behavior near $T_c$ is slightly different in the steepness of the curves due to the difference in the number of sites. The major difference between the two is the actual value of $T_c$, which is slightly higher for Penrose. One potential explanation is that the average Penrose lattice site is connected to more bonds than the average square lattice site. This could lead to a larger effective spin coupling which allows the critical temperature to rive.

The square lattice can be stretched into a rectangular lattice by allowing the horizontal and vertical bonds to differ in length. Assigning different interaction strengths $J_1$ and $J_2$ to the two bond lengths gives a free parameter $J_1/J_2$. If we let $J \propto \ell^{-2}$ where $\ell$ is the bond length, then the phase diagram is further parameterized by the length ratio $a/b$.

Let us define the anisotropy $\eta$ of the lattice as 
\begin{equation}
  \frac{a}{b} = \frac{2}{\pi}\tan^{-1}\eta,\qquad \eta = \tan\frac{\pi a}{2b}.
\end{equation}
We fix the product of the lengths $ab=1$. Then $\eta \in (-1,1)$ and $\eta=0$ implies $a=b$. For the square lattice in particular, rotational symmetry guarantees that the phase diagram $T_c(\eta)$ is symmetric under $a\leftrightarrow b$, or $\eta \rightarrow -\eta$.

Figure \ref{fig:rect} shows the critical temperature as a function of $\eta$ under the Ising, XY, and Heisenberg models. As expected, the result is symmetric around $\eta \rightarrow -\eta$.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../figs/rect-phase.pdf}
  \caption{The phase diagram of the rectangular lattice as a function of the anisotropy $\eta$, where $\eta = 0$ is the square lattice and $\eta\rightarrow \pm 1$ is the anisotropic limit.}
  \label{fig:rect}
\end{figure}

A similar procedure shows the phase diagram for the einstein tiling, where $a$ and $b$ are now the two lengths shown in figure \ref{fig:einstein}.

\section{Transverse Ising Model}
\label{sec:tim}

\section{Conclusion}
\label{sec:conclusion}

\appendix

\section{The Fluctuation-Dissipation Theorem and Magnetic Susceptibility}
\label{app:fluctuation-dissipation}

\section{Equivalence Between the Transverse Ising Model and Classical Ising Model}
\label{app:tim-equivalence}
